---
title: "Time Series Econometrics – Assignment 3"
subtitle: "Model Selection, Forecasting & Dickey-Fuller Testing"
author:
  - "Aadhitya Tejaswin Prakash Sridevi"
  - "Johannes Felix Boppert"
  - "Keno Klebinger"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 4
)
```


```{r packages}

required_pkgs <- c(
  "readxl", "tseries", "tidyverse", "tsibble",
  "fable", "fabletools", "feasts", "urca", "forecast"
)

new_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[, "Package"])]
if (length(new_pkgs)) install.packages(new_pkgs)

library(readxl)
library(tseries)
library(tidyverse)
library(tsibble)
library(fable)
library(fabletools)
library(feasts)
library(urca)
library(forecast)
```

```{r data-loading}
# Load the Singapore visitor arrivals data
visitors_raw <- read_excel("SingaporeVisitors.xlsx")

# Inspect the raw structure
head(visitors_raw)
str(visitors_raw)
```

```{r data-preparation}
# Rename columns for convenience
visitors <- visitors_raw %>%
  rename(
    date     = `Month Year`,
    arrivals = `Visitor Arrivals`
  ) %>%
  mutate(date = as.Date(date))

# Restrict to training period: January 2008 to December 2018
train <- visitors %>%
  filter(date >= as.Date("2008-01-01") & date <= as.Date("2018-12-01"))

# Test period: January 2019 to December 2019
test <- visitors %>%
  filter(date >= as.Date("2019-01-01") & date <= as.Date("2019-12-01"))

# Verify dimensions
cat("Training rows:", nrow(train), "\n")  # expect 132
cat("Test rows:    ", nrow(test),  "\n")  # expect 12

# Convert training data to ts object (frequency = 12 for monthly)
train_ts <- ts(train$arrivals, start = c(2008, 1), frequency = 12)

# Quick plot of raw training series
autoplot(train_ts) +
  labs(title = "Singapore Visitor Arrivals (Jan 2008 – Dec 2018)",
       x = "Year", y = "Arrivals") +
  theme_minimal()
```

The raw training series (January 2008 – December 2018) exhibits two prominent 
features. First, there is a clear **upward trend** — visitor arrivals grow from 
roughly 800,000 per month in 2008 to over 1,500,000 by 2018. Second, regular 
**seasonal fluctuations** are visible throughout, with recurring peaks and troughs 
within each year, consistent with the monthly seasonality ($m = 12$) expected in 
tourism data. 

Both features confirm that the series is non-stationary in its raw 
form, and that differencing will be required before ARMA modelling can proceed.

## Problem 1: Model Selection and Forecasting

### Part (a): Seasonal Differencing

Since the series exhibits a seasonal pattern with period $m = 12$, we apply a 
seasonal difference of lag 12, i.e. $\nabla_{12} y_t = y_t - y_{t-12}$, to 
remove the seasonal component. We then plot the differenced series and examine 
its ACF to assess whether stationarity has been achieved.

```{r seasonal-differencing}
# Apply seasonal differencing (lag 12)
train_ts_sdiff <- diff(train_ts, lag = 12)

# Plot the seasonally differenced series
autoplot(train_ts_sdiff) +
  labs(title = "Seasonally Differenced Visitor Arrivals (Lag 12)",
       x = "Year", y = expression(nabla[12] ~ y[t])) +
  theme_minimal()

# ACF of the seasonally differenced series
ggAcf(train_ts_sdiff, lag.max = 48) +
  labs(title = "ACF of Seasonally Differenced Series") +
  theme_minimal()
```

The seasonally differenced series $\nabla_{12} y_t = y_t - y_{t-12}$ oscillates 
around zero with no visible upward or downward trend, suggesting that the seasonal 
differencing has successfully removed the deterministic seasonal component. However, 
the series does not appear to fluctuate around a constant mean throughout the sample 
— there is a notable downward drift between 2013 and 2015 followed by a recovery — 
which raises some concern about full stationarity.

The ACF provides further evidence against stationarity: the autocorrelations decay 
very slowly and remain significant for many lags (up to lag 10–11), which is 
characteristic of a unit root process. Additionally, significant spikes reappear 
around lags 36 and 48, suggesting residual long-run dependence. Overall, the 
seasonally differenced series does **not** appear fully stationary, and a further 
non-seasonal difference is likely required.

### Part (b): ADF Test and Non-Seasonal Differencing

To formally assess whether a non-seasonal difference is needed, we apply an 
Augmented Dickey-Fuller (ADF) test to the seasonally differenced series. The null 
hypothesis is that the series contains a unit root (non-stationary); rejection 
of the null indicates stationarity suitable for ARMA modelling.


```{r adf-test}
# ADF test on the seasonally differenced series
adf_result <- adf.test(train_ts_sdiff)
print(adf_result)

# The ADF test p-value guides our decision:
# p > 0.05 -> fail to reject unit root -> apply non-seasonal differencing
# p < 0.05 -> reject unit root -> series is stationary

# Apply non-seasonal differencing since ACF suggested non-stationarity
train_ts_diff <- diff(train_ts_sdiff, lag = 1)

# Plot the doubly differenced series
autoplot(train_ts_diff) +
  labs(title = "Doubly Differenced Visitor Arrivals (Seasonal + Non-Seasonal)",
       x = "Year", y = expression(nabla ~ nabla[12] ~ y[t])) +
  theme_minimal()

# Confirm stationarity with ADF test on doubly differenced series
adf_result2 <- adf.test(train_ts_diff)
print(adf_result2)
```

The ADF test on the seasonally differenced series yields a Dickey-Fuller statistic 
of $-3.055$ with a p-value of $0.139$, which is well above the 5% significance 
level. We therefore **fail to reject** the null hypothesis of a unit root, 
confirming that the seasonally differenced series is not yet stationary and that 
a further non-seasonal difference is required.

After applying an additional non-seasonal difference, the doubly differenced series 
$\nabla \nabla_{12} y_t$ fluctuates around zero with no visible trend or drift, 
and the ADF test yields a Dickey-Fuller statistic of $-5.431$ with a p-value of 
$0.01$ (the minimum reportable by the test). We **reject** the null hypothesis of 
a unit root at all conventional significance levels. The doubly differenced series 
is therefore stationary and suitable for ARMA modelling. This corresponds to a 
SARIMA specification with $d = 1$ and $D = 1$.


### Part (c): Box-Jenkins Approach — ACF and PACF Inspection

With the doubly differenced series $\nabla \nabla_{12} y_t$ confirmed as 
stationary, we follow the Box-Jenkins approach and inspect its ACF and PACF 
to identify candidate ARMA$(p, q)$ orders. We also examine whether any residual 
seasonality remains at the seasonal lags (multiples of 12).

```{r acf-pacf}
# ACF of the doubly differenced series
p1 <- ggAcf(train_ts_diff, lag.max = 48) +
  labs(title = "ACF of Doubly Differenced Series") +
  theme_minimal()

# PACF of the doubly differenced series
p2 <- ggPacf(train_ts_diff, lag.max = 48) +
  labs(title = "PACF of Doubly Differenced Series") +
  theme_minimal()

# Display side by side
library(gridExtra)
grid.arrange(p1, p2, ncol = 2)
```

The ACF and PACF of the doubly differenced series $\nabla \nabla_{12} y_t$ reveal 
the following patterns:

**Non-seasonal lags (lags 1–11):** The ACF shows a significant spike at lag 1 and 
a few marginally significant spikes at lags 2–3, before cutting off. The PACF also 
shows a significant spike at lag 1 and then dies off gradually. This pattern is 
broadly consistent with a low-order MA(1) or ARMA(1,1) process for the non-seasonal 
component.

**Seasonal lags (multiples of 12):** The ACF shows significant spikes at lags 12, 
24, and 36, suggesting residual seasonal autocorrelation that was not fully removed 
by the seasonal difference alone. The PACF similarly shows significant spikes at 
seasonal lags. This indicates that seasonal AR or MA terms may be needed in the 
final SARIMA specification.

Based on this inspection, candidate non-seasonal orders are $p, q \in \{0, 1, 2, 3\}$, 
which will be compared formally using information criteria in Part (d).


### Part (d): ARMA Model Selection via AIC and BIC

We now systematically compare all ARMA$(p, q)$ models for $0 \leq p, q \leq 3$ 
fitted to the doubly differenced series, using AIC and BIC as selection criteria. 
The preferred model(s) are then estimated and subjected to residual diagnostics 
to verify adequacy.

```{r model-selection}
# Grid search over ARMA(p,q) for 0 <= p,q <= 3
# Fitted to the doubly differenced series

p_orders <- 0:3
q_orders <- 0:3

results <- data.frame(
  p     = integer(),
  q     = integer(),
  AIC   = numeric(),
  BIC   = numeric()
)

for (p in p_orders) {
  for (q in q_orders) {
    tryCatch({
      fit <- Arima(train_ts_diff, order = c(p, 0, q))
      results <- rbind(results, data.frame(
        p   = p,
        q   = q,
        AIC = AIC(fit),
        BIC = BIC(fit)
      ))
    }, error = function(e) NULL)
  }
}

# Sort by AIC and BIC
results_aic <- results[order(results$AIC), ]
results_bic <- results[order(results$BIC), ]

cat("--- Top 5 models by AIC ---\n")
print(head(results_aic, 5))

cat("\n--- Top 5 models by BIC ---\n")
print(head(results_bic, 5))

# Highlight best models
cat("\nBest model by AIC: ARMA(", results_aic$p[1], ",", results_aic$q[1], ")\n")
cat("Best model by BIC: ARMA(", results_bic$p[1], ",", results_bic$q[1], ")\n")
```

The grid search over all ARMA$(p,q)$ models with $0 \leq p, q \leq 3$ yields the 
following conclusions:

**AIC criterion:** The preferred model is ARMA$(2, 1)$ with AIC $= 2915.77$, 
followed closely by ARMA$(1, 2)$ (AIC $= 2916.20$) and ARMA$(3, 2)$ 
(AIC $= 2917.10$). The differences among the top models are small, suggesting 
that several specifications fit the data similarly well.

**BIC criterion:** The preferred model is ARMA$(0, 1)$ with BIC $= 2926.93$, 
followed by ARMA$(1, 0)$ (BIC $= 2928.36$). BIC applies a heavier penalty for 
additional parameters than AIC, and therefore favours the more parsimonious 
specifications.

We proceed with two candidate models: **ARMA$(2,1)$** as selected by AIC, and 
**ARMA$(0,1)$** as selected by BIC. Both will be estimated on the doubly 
differenced series and subjected to residual diagnostics. In the SARIMA 
framework, these correspond to ARIMA$(2,1,1)(0,1,0)_{12}$ and 
ARIMA$(0,1,1)(0,1,0)_{12}$ respectively.

#### Estimation and Residual Diagnostics

We now estimate both selected models and perform residual diagnostics to verify 
that the residuals behave as white noise — a necessary condition for the model 
to be adequate. We inspect the residual ACF and apply the Ljung-Box test for 
remaining serial correlation.

```{r residual-diagnostics}
# Estimate the two selected models on the ORIGINAL training series
fit_aic <- Arima(train_ts, order = c(2, 1, 1),
                 seasonal = list(order = c(0, 1, 0), period = 12))
fit_bic <- Arima(train_ts, order = c(0, 1, 1),
                 seasonal = list(order = c(0, 1, 0), period = 12))

cat("=== AIC-preferred: ARIMA(2,1,1)(0,1,0)[12] ===\n")
print(summary(fit_aic))
cat("\n=== BIC-preferred: ARIMA(0,1,1)(0,1,0)[12] ===\n")
print(summary(fit_bic))

# --- Residual diagnostics for ARIMA(2,1,1) ---
res_aic <- residuals(fit_aic)
res_bic <- residuals(fit_bic)

cat("\n--- Ljung-Box Test: ARIMA(2,1,1)(0,1,0)[12] ---\n")
print(Box.test(res_aic, lag = 20, type = "Ljung-Box", fitdf = 3))

cat("\n--- Jarque-Bera Test: ARIMA(2,1,1)(0,1,0)[12] ---\n")
print(jarque.bera.test(res_aic))

cat("\n--- Ljung-Box Test: ARIMA(0,1,1)(0,1,0)[12] ---\n")
print(Box.test(res_bic, lag = 20, type = "Ljung-Box", fitdf = 1))

cat("\n--- Jarque-Bera Test: ARIMA(0,1,1)(0,1,0)[12] ---\n")
print(jarque.bera.test(res_bic))

# ACF of residuals
p1 <- ggAcf(res_aic, lag.max = 48) +
  labs(title = "ACF of Residuals: ARIMA(2,1,1)(0,1,0)[12]") +
  theme_minimal()
p2 <- ggAcf(res_bic, lag.max = 48) +
  labs(title = "ACF of Residuals: ARIMA(0,1,1)(0,1,0)[12]") +
  theme_minimal()

# ACF of squared residuals (heteroskedasticity check)
p3 <- ggAcf(res_aic^2, lag.max = 48) +
  labs(title = "ACF of Squared Residuals: ARIMA(2,1,1)(0,1,0)[12]") +
  theme_minimal()
p4 <- ggAcf(res_bic^2, lag.max = 48) +
  labs(title = "ACF of Squared Residuals: ARIMA(0,1,1)(0,1,0)[12]") +
  theme_minimal()

grid.arrange(p1, p2, p3, p4, ncol = 2)
```
**Model Estimates:**

The AIC-preferred ARIMA$(2,1,1)(0,1,0)_{12}$ yields coefficients 
$\hat{\phi}_1 = -1.229$, $\hat{\phi}_2 = -0.351$, $\hat{\theta}_1 = 1.000$, 
with $\hat{\sigma}^2 = 2.367 \times 10^9$ and a training MAPE of $2.97\%$. 
The BIC-preferred ARIMA$(0,1,1)(0,1,0)_{12}$ is considerably more parsimonious, 
with a single coefficient $\hat{\theta}_1 = -0.329$, $\hat{\sigma}^2 = 2.518 \times 10^9$, 
and a training MAPE of $3.04\%$.

**Residual Diagnostics:**

Both models pass all three diagnostic tests:

- **Serial correlation (Ljung-Box):** ARIMA$(2,1,1)$ yields $Q(20) = 21.75$ 
(p-value $= 0.195$) and ARIMA$(0,1,1)$ yields $Q(20) = 25.06$ (p-value $= 0.159$). 
We fail to reject the null of no serial correlation in both cases.

- **Normality (Jarque-Bera):** ARIMA$(2,1,1)$ yields $JB = 1.826$ (p-value $= 0.401$) 
and ARIMA$(0,1,1)$ yields $JB = 0.932$ (p-value $= 0.627$). Residuals are 
consistent with normality in both cases.

- **Heteroskedasticity (ACF of squared residuals):** Both models show all spikes 
within the confidence bands, providing no evidence of remaining conditional 
heteroskedasticity.

Despite passing diagnostics, the residual ACF plots still show significant spikes 
at seasonal lags (particularly lags 12 and 24), indicating that neither model 
fully captures the seasonal dynamics of the series.

### Part (e): Automatic SARIMA Selection

Since the models in Part (d) lack explicit seasonal AR and MA components and 
exhibit residual seasonal autocorrelation, we now use an automatic procedure 
via the \texttt{fable} package to search over SARIMA specifications with 
seasonal orders $0 \leq P, Q \leq 2$ and $D = 1$, alongside non-seasonal 
orders $0 \leq p, q \leq 3$ and $d = 1$. The automatically selected model 
is then compared against the Part (d) candidates.

```{r sarima-auto}
# Convert training data to tsibble format for fable
train_tsibble <- train %>%
  mutate(date = yearmonth(date)) %>%
  as_tsibble(index = date)

# Automatic SARIMA search over specified orders
fit_auto <- train_tsibble %>%
  model(
    ARIMA = ARIMA(arrivals ~ pdq(0:3, 1, 0:3) + PDQ(0:2, 1, 0:2))
  )

# Report the selected model
cat("=== Automatically Selected SARIMA Model ===\n")
report(fit_auto)

# Extract coefficients and compute t-statistics
coef_auto <- fit_auto %>%
  tidy() %>%
  select(term, estimate, std.error) %>%
  mutate(t_statistic = round(estimate / std.error, 2))

cat("\n=== Coefficient t-Statistics: Auto SARIMA ===\n")
print(coef_auto)

# Compare with Part (d) models — refit in fable for consistency
fit_compare <- train_tsibble %>%
  model(
    AIC_model   = ARIMA(arrivals ~ pdq(2, 1, 1) + PDQ(0, 1, 0)),
    BIC_model   = ARIMA(arrivals ~ pdq(0, 1, 1) + PDQ(0, 1, 0)),
    Auto_SARIMA = ARIMA(arrivals ~ pdq(0:3, 1, 0:3) + PDQ(0:2, 1, 0:2))
  )

# Print AIC and BIC for all three
cat("\n=== Model Comparison: AIC and BIC ===\n")
glance(fit_compare) %>%
  select(.model, AIC, BIC) %>%
  arrange(AIC) %>%
  print()
```

The automatic procedure selects **ARIMA$(1,1,2)(1,1,1)_{12}$** as the optimal 
SARIMA model. The estimated coefficients and their t-statistics are:

| Coefficient | Estimate | Std. Error | t-statistic |
|-------------|----------|------------|-------------|
| $\hat{\phi}_1$ (ar1) | $-0.850$ | $0.087$ | $-9.75$ |
| $\hat{\theta}_1$ (ma1) | $0.570$ | $0.125$ | $4.56$ |
| $\hat{\theta}_2$ (ma2) | $-0.396$ | $0.091$ | $-4.35$ |
| $\hat{\Phi}_1$ (sar1) | $0.445$ | $0.225$ | $1.98$ |
| $\hat{\Theta}_1$ (sma1) | $-0.752$ | $0.194$ | $-3.88$ |

Most coefficients are clearly statistically significant. The seasonal AR 
coefficient $\hat{\Phi}_1 = 0.445$ is only marginally significant 
(t-statistic $= 1.98$), but is retained as the overall model is selected 
by the automatic procedure on AIC grounds.

The three-way model comparison reveals a clear ranking:

| Model | AIC | BIC |
|-------|-----|-----|
| ARIMA$(1,1,2)(1,1,1)_{12}$ | **2908.04** | 2924.72 |
| ARIMA$(2,1,1)(0,1,0)_{12}$ | 2913.91 | 2925.03 |
| ARIMA$(0,1,1)(0,1,0)_{12}$ | 2916.76 | **2922.31** |

The automatically selected SARIMA model achieves the lowest AIC by nearly 
6 points over the AIC-preferred Part (d) model, confirming that the inclusion 
of seasonal AR and MA terms $(\Phi_1, \Theta_1)$ substantially improves fit. 
The BIC-preferred model remains ARIMA$(0,1,1)(0,1,0)_{12}$ due to its extreme 
parsimony, but the automatic SARIMA is competitive on BIC as well.

### Part (f): Forecasting and Forecast Evaluation

We now generate 12-step ahead forecasts for the test period January 2019 to 
December 2019 using all three model specifications. Forecast accuracy is evaluated 
using three metrics: Root Mean Squared Error (RMSE), Mean Absolute Error (MAE), 
and Mean Absolute Percentage Error (MAPE).

```{r forecasting}
# Convert test data to tsibble format
test_tsibble <- test %>%
  mutate(date = yearmonth(date)) %>%
  as_tsibble(index = date)

# Generate 12-step ahead forecasts for all three models
forecasts <- fit_compare %>%
  forecast(h = 12)

# Plot forecasts against actual test data
forecasts %>%
  autoplot(train_tsibble, level = 95) +
  autolayer(test_tsibble, arrivals, colour = "black", linewidth = 0.8) +
  labs(title = "12-Month Ahead Forecasts vs Actual (Jan–Dec 2019)",
       x = "Month", y = "Visitor Arrivals") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Compute accuracy metrics against test set
accuracy_metrics <- forecasts %>%
  accuracy(test_tsibble)

# Display RMSE, MAE, MAPE for each model
cat("=== Forecast Accuracy: Jan–Dec 2019 ===\n")
accuracy_metrics %>%
  select(.model, RMSE, MAE, MAPE) %>%
  arrange(RMSE) %>%
  print()
```
The forecast plot shows that all three models track the general level and seasonal 
pattern of the test period well, with the 95\% prediction intervals comfortably 
containing the actual observations throughout January–December 2019.

The accuracy metrics confirm a clear ranking:

| Model | RMSE | MAE | MAPE |
|-------|------|-----|------|
| ARIMA$(1,1,2)(1,1,1)_{12}$ | **40,397** | **34,066** | **2.12\%** |
| ARIMA$(0,1,1)(0,1,0)_{12}$ | 45,695 | 37,710 | 2.35\% |
| ARIMA$(2,1,1)(0,1,0)_{12}$ | 45,950 | 37,388 | 2.33\% |

The automatically selected SARIMA$(1,1,2)(1,1,1)_{12}$ outperforms both Part (d) 
models on all three metrics by a substantial margin. Its MAPE of $2.12\%$ indicates 
that forecasts deviate from actual arrivals by just over 2\% on average — a strong 
result for a 12-month horizon. The inclusion of seasonal AR and MA components 
$(\Phi_1, \Theta_1)$ is clearly beneficial, capturing the seasonal dynamics that 
the simpler Part (d) specifications miss. The two Part (d) models perform similarly 
to each other, with the AIC model marginally better on MAE and the BIC model 
marginally better on RMSE.

## Problem 2: Dickey-Fuller Testing via Monte Carlo Simulation

In this problem we investigate the finite-sample distribution of the 
Augmented Dickey-Fuller (ADF) t-statistic through Monte Carlo simulation. 
The ADF test is programmed from scratch — no existing R functions for the 
ADF test are used. We simulate data from the AR(2) process
$$y_t = \alpha_1 y_{t-1} + \alpha_2 y_{t-2} + \varepsilon_t, \quad 
\varepsilon_t \sim \text{iid}\, \mathcal{N}(0, \sigma^2),$$
with $\alpha_1 = 0.7$, $\alpha_2 = 0.3$, $\sigma = 0.9$, and initial 
conditions $y_{-1} = y_0 = 0$.

### Part (a)–(d): Monte Carlo Setup

We repeat the simulation for $T \in \{25, 100, 500, 5000\}$ with $M = 1000$ 
replications each. For each simulated series we estimate three ADF regression 
specifications and record the t-statistic on $\phi$:

- **(i)** No constant, no trend: $\Delta y_t = \phi y_{t-1} + \alpha_1^* \Delta y_{t-1} + \varepsilon_t$
- **(ii)** With constant: $\Delta y_t = c + \phi y_{t-1} + \alpha_1^* \Delta y_{t-1} + \varepsilon_t$  
- **(iii)** With constant and trend: $\Delta y_t = c + \delta t + \phi y_{t-1} + \alpha_1^* \Delta y_{t-1} + \varepsilon_t$


```{r monte-carlo-setup}
set.seed(2026)

alpha1 <- 0.7
alpha2 <- 0.3
sigma  <- 0.9
M      <- 1000
T_vals <- c(25, 100, 500, 5000)

# Simulate AR(2) DGP: y_t = alpha1*y_{t-1} + alpha2*y_{t-2} + eps_t
# y_{-1} = y_0 = 0
simulate_ar2 <- function(T, alpha1, alpha2, sigma) {
  y <- numeric(T + 2)
  y[1] <- 0
  y[2] <- 0
  eps <- rnorm(T, mean = 0, sd = sigma)
  for (t in 1:T) {
    y[t + 2] <- alpha1 * y[t + 1] + alpha2 * y[t] + eps[t]
  }
  return(y[3:(T + 2)])
}

# ADF regression: reparameterised as
# Delta(y_t) = phi*y_{t-1} + alpha1_star*Delta(y_{t-1}) + [c] + [delta*t] + eps_t
# phi = alpha1 + alpha2 - 1 (unit root parameter)
# We record the t-statistic on phi (coefficient on y_{t-1})

adf_tstat <- function(y, spec) {
  T   <- length(y)
  dy  <- diff(y)              # Delta y_t, length T-1
  yl  <- y[-T]                # y_{t-1},   length T-1
  dyl <- c(NA, dy[-( T-1)])  # Delta y_{t-1}, length T-1 (first entry NA)

  # Build data frame and drop first row (NA in dyl)
  df <- data.frame(
    dy  = dy,
    yl  = yl,
    dyl = dyl,
    tt  = 2:T
  )
  df <- df[complete.cases(df), ]   # removes first row where dyl is NA

  # Fit regression by spec
  if (spec == "none") {
    fit <- lm(dy ~ yl + dyl - 1, data = df)
  } else if (spec == "const") {
    fit <- lm(dy ~ yl + dyl, data = df)
  } else if (spec == "trend") {
    fit <- lm(dy ~ yl + dyl + tt, data = df)
  }

  # Extract t-statistic on yl (always named "yl" in lm output)
  coefs <- summary(fit)$coefficients
  return(coefs["yl", "t value"])
}

cat("Setup complete. Functions defined.\n")
cat("T values:", T_vals, "\n")
cat("M replications:", M, "\n")

# Quick sanity check on one series
set.seed(2026)
y_test <- simulate_ar2(100, alpha1, alpha2, sigma)
cat("\nSanity check t-stats (T=100, one series):\n")
cat("none: ",  round(adf_tstat(y_test, "none"),  3), "\n")
cat("const:",  round(adf_tstat(y_test, "const"), 3), "\n")
cat("trend:",  round(adf_tstat(y_test, "trend"), 3), "\n")
```

```{r monte-carlo-run}
# Storage: list indexed by T, then by spec
specs <- c("none", "const", "trend")

# Run simulation
results_mc <- list()

for (T in T_vals) {
  tstats <- matrix(NA, nrow = M, ncol = 3)
  colnames(tstats) <- specs
  
  for (m in 1:M) {
    y <- simulate_ar2(T, alpha1, alpha2, sigma)
    for (s in specs) {
      tstats[m, s] <- tryCatch(
        adf_tstat(y, s),
        error = function(e) NA
      )
    }
  }
  results_mc[[as.character(T)]] <- tstats
  cat("T =", T, "done.\n")
}

cat("\nSimulation complete.\n")
```
```{r quantile-tables}
# Quantile levels required by the assignment
probs <- c(0.01, 0.025, 0.05, 0.10, 0.90, 0.95, 0.975, 0.99)

# Asymptotic ADF critical values for reference
# Source: MacKinnon (1996)
# For H0: unit root, left-tail critical values at 1%, 5%, 10%
adf_cv <- list(
  none  = c(-2.56, -1.94, -1.62),   # no constant, no trend
  const = c(-3.43, -2.86, -2.57),   # constant
  trend = c(-3.96, -3.41, -3.13)    # constant + trend
)

# Function to build quantile table for one spec
make_table <- function(spec) {
  tab <- matrix(NA, nrow = length(T_vals), ncol = length(probs))
  rownames(tab) <- paste0("T=", T_vals)
  colnames(tab) <- probs
  for (i in seq_along(T_vals)) {
    T <- T_vals[i]
    tstats <- results_mc[[as.character(T)]][, spec]
    tab[i, ] <- quantile(tstats, probs = probs, na.rm = TRUE)
  }
  return(tab)
}

# Print tables for all three specs
for (spec in specs) {
  cat("\n=== Spec (", switch(spec,
    none  = "i) No constant, no trend",
    const = "ii) Constant",
    trend = "iii) Constant + Trend"
  ), ") ===\n")
  tab <- make_table(spec)
  print(round(tab, 3))
  cat("\nAsymptotic ADF critical values (1%, 5%, 10%):",
      adf_cv[[spec]], "\n")
}
```

```{r density-plots}
# Extract T=5000 t-statistics for all three specs
tstats_5000 <- as.data.frame(results_mc[["5000"]])
colnames(tstats_5000) <- c("None", "Constant", "Trend")

# Add standard normal draws for comparison
set.seed(2026)
tstats_5000$Normal <- rnorm(M)

# Reshape to long format
tstats_long <- tstats_5000 %>%
  pivot_longer(cols = everything(),
               names_to  = "Specification",
               values_to = "t_stat") %>%
  mutate(Specification = factor(Specification,
    levels = c("None", "Constant", "Trend", "Normal")))

# Plot densities
ggplot(tstats_long, aes(x = t_stat, colour = Specification,
                         fill = Specification)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  scale_colour_manual(
    values = c("None"     = "#E41A1C",
               "Constant" = "#377EB8",
               "Trend"    = "#4DAF4A",
               "Normal"   = "black"),
    labels = c("(i) No constant/trend",
               "(ii) Constant",
               "(iii) Constant + Trend",
               "N(0,1)")
  ) +
  scale_fill_manual(
    values = c("None"     = "#E41A1C",
               "Constant" = "#377EB8",
               "Trend"    = "#4DAF4A",
               "Normal"   = "black"),
    labels = c("(i) No constant/trend",
               "(ii) Constant",
               "(iii) Constant + Trend",
               "N(0,1)")
  ) +
  xlim(-10, 8) +
  labs(title    = "ADF t-Statistic Distributions vs N(0,1): T = 5000",
       x        = "t-statistic",
       y        = "Density",
       colour   = "Specification",
       fill     = "Specification") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**Quantile Tables — Discussion:**

The simulated quantile tables reveal several important features of the 
finite-sample ADF t-statistic distributions:

**Unit root in the DGP:** The data generating process has 
$\alpha_1 + \alpha_2 = 0.7 + 0.3 = 1$. The AR polynomial 
$1 - 0.7L - 0.3L^2$ factors as $(1-L)(1+0.3L)$, so the process contains 
a unit root. This is why the t-statistics for $\phi$ follow a 
Dickey-Fuller type distribution rather than a standard normal.

**Left tail (critical values):** For all three specifications, the simulated 
left-tail quantiles are more negative than those of the standard normal 
($-1.645$ at 5\%), confirming that standard normal critical values cannot 
be used for ADF testing. For example, under specification (i) at $T = 5000$, 
the 5\% quantile is $-1.940$, close to the MacKinnon asymptotic value of 
$-1.94$.

**Convergence with $T$:** As $T$ increases, the simulated quantiles converge 
toward the asymptotic ADF critical values of MacKinnon (1996). Convergence 
is already quite good at $T = 500$ for all three specifications.

**Finite-sample distortions:** At small $T$ (especially $T = 25$), the 
distributions are more dispersed and the quantiles deviate more from their 
asymptotic counterparts, confirming that the ADF test has weaker size 
properties in small samples.

**Adding deterministic components shifts the distribution left:** Each 
additional deterministic term (constant, trend) shifts the distribution 
further to the left, requiring more negative critical values to achieve 
the same significance level.

**Density plot — $T = 5000$:** The density plot confirms that all three 
ADF t-statistic distributions are shifted to the left relative to 
$\mathcal{N}(0,1)$. The no-constant specification (i) is closest to the 
normal but still distinctly left-shifted. The constant-and-trend 
specification (iii) is the most displaced, with its mass concentrated 
around $-3$ to $-4$. Using standard normal critical values for ADF 
testing would therefore lead to under-rejection of the null hypothesis 
of a unit root.










































